% Included from both -slides and -handout versions.
\documentclass[pdftex]{beamer} % used to trigger beamer mode in Emacs,
                               % normally commented out.

\usetheme{metropolis}

\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{graphicx}
\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage{fancyvrb}
\usepackage{listings}
\begin{document}
\lstset{language=C, escapeinside={(*@}{@*)}, numbers=left,
  basicstyle=\tiny, showspaces=false, showtabs=false}

\title{Introduction to Operating Systems}
\subtitle{Through tracing, analysis, and experimentation}
%\institute{University of Cambridge}
\author{George V. Neville-Neil}
%\author{Dr Robert N. M. Watson}
\date{1 August 2016}

\begin{frame}
  \titlepage
\end{frame}

\section{Storing Information}
\label{sec:storage}

\begin{frame}
  \frametitle{What is a Filesystem?}
  \begin{itemize}
  \item A system for organizing data
  \item Off-line storage
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Some example filesystems}
  \begin{description}
  \item[UFS2] FreeBSD
    \pause
    
  \item[EXT4] Linux
    
    \pause
    
  \item[NTFS] Windows

    \pause

  \item[FAT] Simplest and most common in the world

    \pause

  \item[ZFS] OpenSolaris/Illumos/FreeBSD
  \end{description}
\end{frame}

\begin{frame}
  \frametitle{Filesystem Goals}
  \begin{description}
  \item [Fast] Storage and Retrieval of User Data

    \pause

  \item [Scalable] Files, Data, Disks...

    \pause
    
  \item[Reliable] ``Destroy data once and no one trusts you again'' -
    Kirk McKusick
  \end{description}
\end{frame}

\begin{frame}
  \frametitle{Glossary of Filesystem Terms}
  \begin{description}
  \item[Block] Smallest piece of stored data
    \pause
  \item[File] A collection of blocks
    \pause
  \item [Directory] A collection of files.
  \end{description}
\end{frame}

\begin{frame}[fragile]
  \frametitle{How do programs access stored data?}
  \begin{description}
  \item[syscall] \verb|open|, \verb|close|, \verb|read|, \verb|write|
  \item[filesystem] vnode operations
    \item[physical media] GEOM layer
  \end{description}
\end{frame}

\begin{frame}
  \frametitle{The Filesystems of FreeBSD}
  \begin{description}
  \item [UFS2]
    \begin{itemize}
    \item Fast
    \item Small
    \item Snapshots
    \item Some use of journaling
    \end{itemize}
  \item [ZFS]
    \begin{itemize}
    \item Large storage
    \item Copy on Write
    \item Snapshots
    \end{itemize}
  \end{description}
\end{frame}

\begin{frame}
  \frametitle{Important Parameters}
  \begin{description}
  \item[Block Size] How many bytes in a block?
  \item [Indirect Blocks] How many blocks can we reference?
  \end{description}
\end{frame}


\begin{frame}
  \frametitle{open() a walkthrough}
  \begin{itemize}
  \item What happens when a user opens a file named ``foo''
  \item The open() system call is executed.
  \item The file name is looked up.
  \item The user's permissions are checked against the file's
    permissions.
  \item Notations are made in the kernel to indicate the file is open.
  \item A per-process file descriptor is allocated and returned to the
    caller.
  \item No data is moved from secondary storage until later.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{What does close() do?}
  \begin{itemize}
  \item Remove the file descriptor from the process.
    \begin{itemize}
    \item It can now be reused!
    \end{itemize}
  \item Tells the kernel it can free resources.
    \begin{itemize}
    \item But only if absolutely necessary.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{File Management Operations}
  \begin{description}
  \item[Create] Make a new file.
  \item[Delete] Remove a file.
  \item[Rename] Change the name of a file.
  \end{description}
\end{frame}

\begin{frame}
  \frametitle{Renaming a File}
  \begin{itemize}
  \item The most complex filesystem operation
  \item Idempotent directory updates
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Interactions with Other Subsystems}
  \begin{description}
  \item[Virtual Memory] VNODES vs. VM Objects
  \item[Buffer Cache] Between VM and the Filesystem
  \item[Naming] Name cache
  \item[Networking] Network Filesystem (NFS)
  \end{description}
\end{frame}

\subsection{Naming of Names}
\label{sec:naming}

\begin{frame}[fragile]
  \frametitle{Naming}
  \begin{itemize}
  \item Translate a human name to a location of stored data
  \item \verb+namei+ is the main interface
  \item Names are not stable after translation
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Name Lookup}
  \begin{itemize}
  \item What names are being looked up?
  \end{itemize}
\begin{lstlisting}
dtrace -n 'vfs:namei:lookup:entry { printf("%s", stringof(arg1));}'
CPU     ID                    FUNCTION:NAME
  2  27847                     lookup:entry /bin/ls
  2  27847                     lookup:entry /libexec/ld-elf.so.1
  2  27847                     lookup:entry /etc
  2  27847                     lookup:entry /etc/libmap.conf
  2  27847                     lookup:entry /etc/libmap.conf
\end{lstlisting}
\end{frame}

\begin{frame}
  \frametitle{Name Cache}
  \begin{itemize}
  \item Speeds up searching
  \item Maintains positive and negative results
  \item Invalidation on changes in directories
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Who is missing the cache?}
\begin{lstlisting}
dtrace -n 'vfs:namecache:lookup:miss { printf("%s", stringof(arg1));}'
\end{lstlisting}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Name Cache Module}
  \begin{description}[labelwidth=\widthof{lookup:hit-negative}]
  \item[enter] Add a positive entry
  \item[enter\_negative] Add a negative entry
  \item[lookup:hit] Name found in positive cache
  \item[lookup:hit-negative] Name found in negative cache
  \item[lookup:miss] Name not found in cache
  \item[purge] Remove positive entry
  \item[purge\_negative] Remove negative entry
  \item[zap] Remove positive entry with or without vnode
  \item[zap\_negative] Remove negative entry with or without vnode
  \end{description}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Adding negative entries}
\begin{lstlisting}
 dtrace -n 'vfs:namecache:enter_negative: { printf("%s", stringof(arg1)); }'
\end{lstlisting}
\end{frame}


\subsection{Organizing Data}
\label{sec:organization}

\begin{frame}
  \frametitle{Directories and Files}
  \begin{description}
  \item [Directories] Contain files
  \item [Files] Contain data
  \end{description}
\end{frame}

\begin{frame}
  \frametitle{The VNODE Data Structure}
  \begin{itemize}
  \item The central structure for all storage
  \item Many things have an associated vnode
    \begin{itemize}
    \item Files
    \item Directories
    \item Block Devices
    \item Character Devices
    \item Some Sockets
    \item FIFOs
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{VNODE Operations}
  \begin{itemize}
  \item Classic, kernel OO structure
  \item After a path or name is looked up
  \item Do something with a vnode
  \item \verb|open|, \verb|close|, \verb|read|, \verb|write|
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Visualizing VNODE Operations}
  
\end{frame}

\subsection{Data on Disk}
\label{sec:disk}

\begin{frame}
  \frametitle{GEOM Overview}
  \begin{itemize}
  \item Framework for transforming storage requests
  \item Object Oriented
  \item Shuttles I/O requests between filesystems and devices
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{GEOM Layers}
  \begin{description}
  \item[MBR] Master Boot Record
  \item[BSD] BSD Slice
  \item[ELI] Encryption
  \item[MIRROR] Disk mirroring
  \item[JOURNAL] Journaling
  \item[RAID] Software RAID
\end{description}
\end{frame}

\begin{frame}
  \frametitle{A Visual Example}
  
\end{frame}

\begin{frame}
  \frametitle{Performing I/O}
sudo dtrace -n ::g\_io\*:entry  

This gets us the most likely 8 routines we care about.

\end{frame}

\subsection{Storing Data Remotely}
\label{sec:remote_storage}

\begin{frame}
  \frametitle{NFS: The Network Filesystem}
  \begin{itemize}
  \item Developed by Sun Microsystems in the 1980s
  \item Applies UNIX semantics to networked files
  \item Version 2, 3 and 4
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Storage Review}
  
\end{frame}

\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "lecture4-storage"
%%% End:
